### 2014-07-21 Paper_v1.tex
First commit

\documentclass[12pt]{article}
% Load packages
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{setspace}
%\usepackage{longtable}
\usepackage{colortbl}
%\usepackage{palatino,lettrine}
%\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[wide]{sidecap}
%\usepackage[authoryear,round,comma,sort&compress]{natbib}
\usepackage[square,sort,comma,numbers]{natbib}
%\usepackage[authoryear,round]{natbib}
\usepackage{supertabular}
\usepackage{simplemargins}
\usepackage{comment}
\usepackage{lineno}

\urlstyle{rm}

@@

%\usepackage{geometry}
\usepackage{fullpage}

%\bibliographystyle{plain}
\bibliographystyle{plos2009}

@@

%Publication style settings

% Single space'd bib -
\setlength\bibsep{0pt}

\renewcommand{\rmdefault}{phv}\renewcommand{\sfdefault}{phv}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}

@@

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Dynamically Dimensioned Particle Swarm Optimization}}
\vspace{0.05in}
{\par \centering \large{Adithya Sagar, Christine Shoemaker and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering \large{School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Modeling cell free metabolism}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{BMC~Systems~Biology}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
{\par \centering Email: jdv27@cornell.edu} 
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}

{\noindent \textbf{Keywords:}~Parameter identification, Mathematical modeling}

@@

\linenumbers

\section*{Introduction}

\clearpage

\section*{Results}

Coagulation is an archetype biochemical network that is highly interconnected and tightly regulated with multiple positive and negative feedback loops. 
The biochemistry underlying coagulation has been well studied [REFHERE], and reliable experimental coagulation models have been developed [REFHERE].

This makes it an ideal system for mathematical modeling and parameter estimation. Coagulation is regulated by a set of serine proteases also known as coagulation factors and blood platelets. The coagulation factors are generally in an inactive state and are known as zymogens. These zymogens are activated when through certain triggers. These trigger events like injury or trauma or sepsis exposure factors like collagen, tissue factor and von Willebrand factor (vWF) to blood. The exposure of these factors to blood kick-starts a series of convergent cascades that lead to conversion of zymogen prothrombinase to thrombin.   Luan et al. modeled coagulation using coupled non-linear ordinary differential equations with 301 reactions and 192 species.

\clearpage

\section*{Discussion}

\clearpage

\section*{Materials and Methods}

\subsection*{Formulation and solution of coagulation model equations.}
\subsection*{Optimization methods.}

\section*{Acknowledgements}
This study was supported by the National Science Foundation GK12 award (DGE-1045513) 
and by the National Science Foundation CAREER award (FILLMEIN).

\clearpage

\bibliography{References_v1}

%\begin{comment}

\clearpage

% Supplemental figures -
% Set the S- 
\renewcommand\thefigure{S\arabic{figure}}
\renewcommand\thetable{T\arabic{table}}
\renewcommand\thepage{S-\arabic{page}}
\renewcommand\theequation{S\arabic{equation}}

% Reset the counters -
\setcounter{equation}{0}
\setcounter{table}{0}
\setcounter{figure}{0}
\setcounter{page}{1}

>>>>

\documentclass[12pt]{article}
% Load packages
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage{multicol}   %Columns

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}

\usepackage[round,numbers,comma,sort&compress]{natbib}
\bibliographystyle{biophysj}

\usepackage{longtable}
\usepackage{supertabular}
\usepackage{setspace}
%\usepackage{lineno}

%*** extra packages ****
\usepackage{colortbl}
\usepackage{color}
\usepackage{simplemargins}

@@

\setallmargins{1in}

@@

%Publication style settings
% Single space'd bib -
\setlength\bibsep{0pt}

\renewcommand{\rmdefault}{phv}\renewcommand{\sfdefault}{phv}

@@

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Physiochemical Modeling and Analysis of the Mammalian Unfolded Protein Response}}
\vspace{0.05in}
{\par \centering \large{Anirikh Chakrabarti and Jeffrey D. Varner$^{*}$}}
\vspace{0.1in}
{\par \centering \large{School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Mammalian unfolded protein response and cellular stress}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Bio Phys J}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Assistant Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
{\par \centering Email: jdv27@cornell.edu} 
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Cellular stressors routinely lead to malfunctions in the folding state of critical proteins, thereby playing a vital role in disease states like cancer, diabetes and cardiovascular ailments. Newly synthesized proteins must fold and assemble into unique three-dimensional structures, in order to become functionally active. Cells monitor protein folding by an inbuilt quality-control system involving both the Endoplasmic Reticulum (ER) and the Golgi apparatus. Incorrectly folded proteins are tagged for degradation via ER associated degradation (ERAD) or sent back through a refolding cycle. However, accumulation of incorrectly folded proteins can trigger a cascade of events, termed the Unfolded Protein Response (UPR). UPR leads to re-establishing the cellular homeostasis. %This is done by increased transcription of chaperones, reduced translation and cellular apoptosis. 
In this study, we developed a family of mechanistic models of eukaryotic UPR, which was composed of a system of ordinary differential equations. The objective of this study was to assemble a series of molecular modules describing different aspects of UPR and subsequently analyze these models for fragility and robustness. Kinetic parameters for these models were estimated by comparing simulations with experimental data. Using POETs and a cross-validation scheme, we developed an ensemble of models, consistent with literature data. Model analysis highlighted the presence of a sequential order for firing of the ER stress transducers in UPR. The counteracting effects of the ER stress transducers converge at feedback regulation of the molecular chaperone BiP. The three main regulators of feedback, ATF4, cleaved ATF6 and XBP1s share the load ensuring the phased response of UPR. However these regulators add on to the overall fragility of the system and allow scope for manipulation of UPR as seen by sensitivity/robustness analysis. Downstream effects of UPR include a balance between cell survival and cell death based on the magnitude of the stress. Model analysis suggested that the cell-death axis was relatively robust owing to redundant routes e.g., APAF-1 dependent and APAF-1 independent routes of apoptosis. However the cell-survival axis was relatively susceptible to perturbations owing to multiple levels of regulation, thereby highlighting importance of key proteins like CHOP.

\vspace{0.1in}
{\noindent \textbf{Keywords:}~ Protein Folding, UPR, PERK, ATF6, IRE1, Apoptosis, Mathematical Modeling, Multi-objective thermal annealing, Sensitivity Analysis.}

@@

\section*{Introduction}
Protein folding is strategically important to cellular function in all organisms. In eukaryotes, secreted, membrane-bound and organelle-targeted proteins are typically processed and folded in the endoplasmic reticulum (ER) \cite{naidoo2009er, ron2002translational, kaufman2002unfolded}. Intracellular perturbations caused by a variety of stressors disturb the specialized environment of the ER leading to the accumulation of misfolded or unfolded proteins \cite{ellgaard2003qce,Fonseca:2009fk}. Shifts in folding capacity have been associated with diseases such as cancer, diabetes and cardiovascular disorders \cite{ellgaard2003qce}. Physiological processes such as aging can also influence protein folding \cite{naidoo2009er}. Normally, cells ensure proper protein folding using a combination of molecular chaperones, foldases and lectins \cite{naidoo2009er}. However, when proper folding can not be restored, unfolded or misfolded proteins are targeted to ER Associated Degradation (ERAD) pathways for processing \cite{kaufman2002unfolded}. If unfolded or misfolded proteins continue to accumulate, eukaryotes induce the unfolded protein response (UPR). In mammalian cells, UPR is a complex signaling program mediated by three ER transmembrane receptors: activating transcription factor 6 (ATF6), inositol requiring kinase 1 (IRE1) and double-stranded RNA-activated protein kinase (PKR)-like endoplasmic reticulum kinase (PERK). UPR performs three functions, adaptation, alarm and apoptosis. During adaptation, the UPR tries to reestablish folding homeostasis by inducing the expression of chaperones that enhance protein folding. Simultaneously, translation is globally attenuated to reduce the ER folding load while the degradation of unfolded proteins is increased. If these steps fail, the UPR induces a cellular alarm and apoptosis program. The alarm phase involves several signal transduction events, ultimately leading to the removal of the translational block and the down-regulation of the expression and activity of pro-survival factors such as the B-cell lymphoma 2 (Bcl2) protein. After the alarm phase, cells can undergo apoptosis, although ER stress can also initiate autophagy \cite{ogata2006aac, yorimitsu2006ers, bernales2006ace, kamimoto2006iic, hoyerhansen2007cmc, kouroku2006esp, fujita2007ter}. Thus, ER folding homeostasis strongly influences mammalian physiology \cite{Fonseca:2009fk}.

In this study we used physiochemical computer modeling as a tool to study the mammalian unfolded protein response. Physiochemical models which describe system-level responses can prioritize experimental directions, generate testable hypothesis and perhaps identify and validate potential therapeutic targets \cite{KITANO2007}. For mammalian UPR, physiochemical modeling might tell us the critical components of each branch, and ultimately what must be manipulated to get a desired network response, for example, enhanced death or survival. However, there are critical limitations with physiochemical models. Foremost amongst these is the large number of unknown model parameters and uncertainty in the model structure. It is typically impossible to uniquely identify model parameters, even with extensive training data and perfect models \cite{Gadkar:2005ad}. Alternatively, ensemble approaches, which use uncertain model families, have emerged as a promising strategy in systems biology and other fields like climate prediction \cite{Battogtokh:2002uq,Kuepfer:2007,Brown:2003qy,Palmer:2005,Song:2009uq}. Their central value has been the ability to quantify simulation uncertainty and to experimentally constrain model predictions. For example, Gutenkunst \emph{et al.} showed that predictions were possible using signal transduction model ensembles, despite sometimes only order of magnitude parameter estimates \cite{Gutenkunst:2007oq}. More recently, Luan \emph{et al.} predicted patient response to therapeutic intervention using an ensemble of human coagulation models where approximately 40\% of the model parameters had a coefficient of variation (CV) greater than 0.5 \cite{Luan:2010fc}. Model ensembles have also been used with parameter dependent analysis techniques, e.g., sensitivity analysis to robustly estimate important signaling network features. For example, Tasseff \emph{et al.} characterized emergent behavior between androgen and growth factor signaling in prostate cancer cell-lines using an ensemble of models and sensitivity analysis \cite{Tasseff:2010rq}. Thus, despite uncertainty, ensembles of physiochemical models can be used to understand qualitative properties of complex biochemical networks.

We developed a population of physiochemical models describing the adaptation, alarm and apoptosis phases of mammalian UPR. Traditionally, it has been hypothesized that there is a sequential order for firing of the ER stress transducers in UPR. The PERK and ATF6 branches are thought to be activated before IRE1 \cite{szegezdi2006mediators} and largely promote ER adaptation to misfolding, while IRE1 is the last one to fire as it has a dual role, transmitting both survival and pro-apoptotic signals. However our modeling analysis suggests that these three branches fire simultaneously with varying rates and the state of the cell in terms of adaptation, alarm or apoptosis is a result of counteracting effects of these three prongs of UPR signaling. The counteracting effects is further substantiated by simulated knockout studies, where in knockout of one ER stress transducer lead to enhancement of the other branches of UPR, e.g, if we want more protein production even with ER stress, PERK KO might suggest as a viable route as that shall lead to no translational block. However the IRE1 branch shall amplify and so will the ATF6 branch and ultimately still lead to death (less compared to WT). So a viable strategy could be to regulate CHOP, as it shall lead to reduced death by regulation of Bcl2. One common playground for the three ER stress transducers is the feedback regulation of BiP expression. Model analysis substantiated the significance of BiP feedback in the overall robustness of the system. Removal of any one branch of BiP feedback lead to increased sensitivity of the other branches. So there was load/responsibility sharing within the system. Interestingly, removal of all nodes of BiP feedback increased the overall robustness of the system. So while BiP feedback is crucial in terms of the cell to adapt to small perturbations, at the same time this makes the system more fragile and susceptible to manipulations. Model analysis further highlighted the presence of redundant routes of regulation of apoptosis such as APAF-1 dependent/independent strategies. While manipulation of pro-death axis of the system (cleaved caspases) was relatively robust to perturbations, manipulation of the pro-survival axis (Bcl2) was relatively feasible via CHOP dependent/independent strategies. Overall UPR was seen to be robust to perturbations (simulated knockout/overexpression studies), thereby highlighting the redundancy and crosstalk within the three branches of UPR. However, BiP regulation at the transcriptional level via intermediates of the ER-stress transducer signaling cascades (e.g. ATF4, XBP1s) was seen to be the key in the regulation of UPR. The EUPR model is available in SBML in the supplemental materials.

\section*{Results}
\subsection*{Formulation of the UPR network architecture.}
The UPR network described the ER folding cycle, ER-associated degradation (ERAD), ER-stress transducer (PERK, IRE1 and ATF6) signaling cascades and stress-induced caspase activation (Fig. \ref{fg:fig1}). The network consisted of 636 protein or mRNA species interconnected by 1090 interactions (Fig. \ref{fg:fig1} Inset). Connectivity was formulated from a comprehensive review of the primary literature  \cite{naidoo2009er, ron2002translational, kaufman2002unfolded,ellgaard2003qce,Fonseca:2009fk,Schroder:2005vn,gotoh:hdc,mccullough2001gsc, belmont2008cga,Hetz:2009qc,Urano:2000uq,szegezdi2006mediators}, and from on-line databases; String-8 \cite{Jensen:2009fk}, NetworKIN \cite{Linding:2007fk} and TRANSFAC. Model connectivity was not specific to a single cell-line. Rather, it was a canonical representation of the pathways involved in monitoring and controlling the folding capacity of a generic well-mixed ER compartment. UPR induction was modeled as the release of BiP from the ER stress transducers, PERK, IRE1$\alpha$ and ATF6 leading initially to adaptation of the folding cycle and then subsequently to alarm and apoptosis. The adaption phase of UPR was marked by general translation attenuation, selective transcriptional programs for key species like bZIP transcription factor ATF4 \cite{lu2004tra}, cellular inhibitor of apoptosis (cIAP) \cite{hamanaka2008pdr}, molecular chaperones e.g., BiP \cite{harding2003isr} and enhanced clearance of accumulated proteins via ERAD. The alarm and apoptosis phases were mediated by the induction of CHOP \cite{ron1992cnd}, regulation of Bcl2, Bcl2-antagonist of cell death (BAD) \cite {wang1999cia} and (TNF) receptor associated factor 2 (TRAF2) \cite{lei2003jpb,putcha2003jmb, yamamoto1999bpa, szegezdi2006mediators} activation (Fig. \ref{fg:fig1}). Model connectivity is available in SBML format from the supplemental materials. Complete details of the interactions in the model are also enlisted in the supplementary materials.

%\clearpage
\subsection*{Estimating a population of canonical UPR models using POETs.}
Despite a significant identifiability challenge, the multiobjective POETs algorithm generated a predictive UPR model population. The three phases of UPR were modeled using mass action kinetics within an ordinary differential equation (ODE) framework. While ODEs and mass-action kinetics are common methods of modeling biological pathways \cite{Schoeberl:2002rv, Chen:2009uq, Helmy:2009kx}, this modeling strategy resulted in a large number of unknown model parameters. These parameters were not uniquely identifiable (data not shown). Instead, we estimated an experimentally constrained population of parameters using multiobjective optimization. A population of the 1726 unknown model parameters (1090 kinetic parameters and 636 initial conditions) was estimated from 33 dynamic and steady state data-sets taken from literature (Table \ref{obj-table}). The residual between model simulations and each of the experimental constraints was simultaneously minimized using the multiobjective POETs algorithm \cite{Song:2010ij}. A leave-eight-out cross-validation strategy was used to independently estimate the training and prediction error over the 33 data sets; we estimated four different model families, where eight of the 33 objectives were reserved for validation and 25 were used model training each family. Thus, each model family was trained and validated on different experimental data. Starting from an initial best-fit initial parameter set (nominal set), more than 25,000 probable models were estimated by POETs from which we selected N = 100 models (25 from each training family) with a Pareto rank of one or less (from approximately 1200 possible choices) for further study. For each objective, the nominal, training (75 models) and prediction (25 models) error was calculated for each objective (Table \ref{obj-table}). Models used for prediction error calculations for a particular objective were \emph{not} trained on that objective. The prediction likelihood was statistically significantly better for 31 of the 33 objective functions at a 95\% confidence level, compared with random parameter sets generated from the nominal set (Table \ref{obj-table}). Thus, POETs generated model families that predicted approximately 94\% of the objective functions with a significantly higher likelihood than a random control. However, the specific value of any given parameter was likely not well described. The coefficient of variation (CV) for the model parameters ranged from 0.5 - 1.6, where approximately 65\% of the parameters were constrained with a CV $\leq$ 1.0 (supplementary materials Fig. S\ref{fg:Para_S1}). The most constrained parameters involved a wide-array of functions e.g., regulation of PERK, eIF2$\alpha$, ATF4, Calcineurin, BiP, CHOP and ATF6 signaling. However, the least constrained parameters involved JNK and apoptosis interactions.

POETs identified Pareto fronts between several objectives, e.g., O13$\times$O14, O25$\times$O29, O11$\times$O29 and O27$\times$O2 in the training data  (Fig. \ref{fg:objective_plot}). Strong Pareto fronts suggested an inability to simultaneously model different aspects of the training data. However, fronts could also result from experimental artifacts, e.g., variation between cell-lines, time-scale differences or from functional relationships in the data. Globally, adaptation and alarm phase training constraints conflicted with those involving apoptosis. For example, objectives involving caspase-7 or caspase-9 activity conflicted with phosphorylated eIF2$\alpha$ levels. Phosphorylation of eIF2$\alpha$ by activated PERK attenuates translation, which decreases the ER folding load. Thus, eIF2$\alpha$ phosphorylation is a key early adaptive event in UPR. On the other hand, caspase-9 is a stress-induced death marker activated only after UPR has failed to restore folding homeostasis. Conflicts between these early and late phase markers suggested the UPR time-scale was perhaps cell-line or perturbation dependent. Similarly, negative feedback may lead to conflicting objectives. For example, XBP1 mRNA measurements (O14) conflicted with CHOP protein measurements (O13), even though these data-sets were taken from the same study and were collected in the same cell-line. XBP1 splicing increased BiP levels, which in turn reduced CHOP protein levels, hence the trade-off. Lastly, in addition to fronts, we also observed strong correlation between objectives. For example, models that performed well for the CHOP protein (O11), also performed well against Procaspase-12 (O22) measurements, even though these were not in the same cell-line or from the same study. Both CHOP and Procaspase-12 are downstream of IRE1/TRAF2/JNK signaling cascade, so these errors were directly correlated.

%\clearpage
\subsection*{The population of UPR models recapitulated adaptation, alarm and apoptotic events across multiple cell-lines and timescales.} 
We trained the UPR ensemble using data from UPR initiation events (e.g., PERK activation) and the downstream activation of proteins involved in apoptosis (BiP, caspase-3 and caspase-7). ER stress induced by exposure to thapsigargin (Tg), a non-competitive inhibitor of SERCA Ca$^{2+}$ transporters, leads to the dissociation of BiP from the ER-stress transducers. For theses initial proof-of-concept simulations, we assumed that the action of Tg and other stress-inducing agents such as dithiothreitol (DTT) was identical, i.e., induction of BiP dissociation. Within 20 min after Tg exposure, PERK was activated and transmitting adaptation signals downstream. The population of UPR models recapitulated the timescale of PERK phosphorylation (Fig. \ref{fg:simulation_comparison}A) as well as its downstream signaling activity, for example, the phosphorylation of eIF2$\alpha$ (Fig. \ref{fg:simulation_comparison}H). The nuclear fraction of ATF4 increased from approximately zero (untreated cells) to a maximum value 4 hrs after Tg exposure. While the model ensemble generally predicted the correct trend, there was significant error in the early time points for ATF4 (Fig. \ref{fg:simulation_comparison}G). The phosphorylation of eIF2$\alpha$ by PERK is required for ATF4 activation. Interestingly, when we compared model simulations of p-eIF2$\alpha$ levels following Tg (1$\mu$M) exposure in MEFs with measurements (O15 \cite{yamamoto2007tim}), the model correctly captured the appropriate behavior. To test the functioning of the ATF6 branch of UPR model, we compared simulations with measurements of cleaved ATF6 in tunicamycin-treated MEFs \cite{lee2002imu}. ER stress leads to the release of the BiP from ATF6. Cleaved ATF6 is then translocated to the nucleus where it up-regulates gene expression \cite{silver1999mtf,hai1989tfa}. Simulations of cleaved ATF6 levels following UPR initiation were consistent with measurements (Fig. \ref{fg:simulation_comparison}C). The formation of p-PERK or p-IRE1 initiates a complex series of events that operate on both short and long time-scales. Signals from the ER stress transducers converge downstream to regulate BiP transcription \cite{malhotra2007era, rao2004mpe,kokame2001iei, yoshida2000aap}. Model ensemble recapitulated correct trends of BiP mRNA with maximum levels $\sim$ 8 hrs, similar to what was seen in experiments done on HEK293 cells (Fig. \ref{fg:simulation_comparison}D) \cite{lin2007ire1}. One of the long-term outcomes of PERK/IRE1 activation is apoptotic cell death. In the proof-of-concept ER-model the link between UPR and apoptosis occurred through the action of eIF2$\alpha$, the dual role of the ATF4 transcription factor and caspase-12 activation by IRE1-TRAF2 signaling axis. We constrained model parameters associated with the activation of the cell-death program using measurements of pro/caspase-7 levels, pro/caspase-9 levels, pro/caspase-3 levels, pro/caspase-12 levels and PARP cleavage mediated by executioner caspases following treatment with 0.5$\mu$M Tg \cite{rao2002cer}. These experiments were performed in Sak2 cells that lack Apaf-1 protein expression \cite{rao2002cer}. Thus, these data allowed us to include a non-Apaf-1 mediated stress-induced caspase activation pathway into the model. The population of models recapitulated caspase-3 (Fig. \ref{fg:simulation_comparison}K), caspase-9 (Fig. \ref{fg:simulation_comparison}J) as well as cleaved Parp levels (Fig. \ref{fg:simulation_comparison}L) following exposure to ER stress-inducers. Interestingly, while PERK activation occurred on the timescale of minutes, initiator and executioner caspase activation occurred over 36 hrs. Thus, the population of UPR models captured complex signaling events occurring across multiple timescales.

\subsection*{Signal flow analysis of UPR highlighted modes of crosstalk and redundancy in BiP and Bcl2 regulation.}
Traditionally, it has been hypothesized that there is a sequential order for firing of the ER stress transducers in UPR. The PERK and ATF6 branches are thought to be activated before IRE1 \cite{szegezdi2006mediators} and largely promote ER adaptation to misfolding, while IRE1 is the last one to fire as it has a dual role, transmitting both survival and pro-apoptotic signals. However our modeling analysis suggests that these three branches fire simultaneously with varying rates and the state of the cell in terms of adaptation, alarm or apoptosis is a result of counteracting effects of these three prongs of UPR signaling (Fig. \ref{fg:proof_of_concept_simulation} and supplementary materials Fig. S\ref{fg:Flux_Supp_Fig1}). The counteracting effects is further substantiated in simulated knockout studies, where in knockout of one ER stress transducer leads to enhancement of the other branches of UPR (supplementary materials Fig. S\ref{fg:Flux_Supp_Fig1}A-C). UPR induction in the model was controlled by manipulation of the generation rate of unfolded or misfolded protein (qP) in the ER compartment. Upon UPR induction, initially ($\leq$ 1hr) the response is damped marking the adaptation phase of UPR. This is followed by an increase in the activity of the IRE1$\alpha$, PERK and ATF6 cascades at $\sim$ 1 hr marking the onset of the alarm phase, which leads to a final steady state $\sim$ 8-10 hrs marking the onset of the commitment or apoptosis phase of UPR (Fig. \ref{fg:proof_of_concept_simulation}). This time frame was consistent with the maximum levels of BiP mRNA upon UPR induction \cite{lin2007ire1}. This analysis was substantiated further by looking at the fluxes at different phases of UPR induction. As compared to P1 (No-UPR Steady State Fig. \ref{fg:proof_of_concept_simulation}), we saw that early on at 1 hr after unfolded protein dose there was a marked increase in ATF4 and CHOP regulation, ATF6 signaling along with unfolded protein sensing and degradation. These are hallmarks of the adaptation-alarm phase of the UPR response (Fig. \ref{fg:cross_plot_flux}D). The apoptosis phase ($\geq$ 8-10 hrs P2), was marked by increased BiP regulation, enhanced ATF4 transcriptional activity, increased mitochondrial membrane permeability and increased apoptotic fluxes. So the cell has committed itself to apoptosis mediated cell death. We compared the ability of the system to recuperate upon reversal of unfolded protein loads from the alarm phase (P4, Fig. \ref{fg:proof_of_concept_simulation} ) and the apoptosis phase (P2, Fig. \ref{fg:proof_of_concept_simulation} ). If we reduced the protein load after the cell was committed to apoptosis (as in P3), we found that the cell continued to function similar to the UPR state even upon protein load reduction after 25 hrs (Fig. \ref{fg:cross_plot_flux}C). Certain modules were seen to reduce like IRE1-TRAF2 signaling, ASK1 activation. However not much difference was seen in terms of apoptotic fluxes, denoting the commitment of the cell to death and to a point of no return. On the contrary if we reduced the load of proteins in the adaptation-alarm phase (P4), we saw that the cell could recuperate using its ERAD machinery and the regulation of BiP (Fig. \ref{fg:cross_plot_flux}E-F).

We further investigated the effect of simulated knockout (KO) and overexpression (OX) of key proteins on the UPR system. Interestingly, PERK and ATF4 KO studies revealed a slower and lower amount of BiP production ($\sim$ 50\%) as compared to WT. However, ATF6 or IRE1 KO did not affect BiP regulation as compared to WT. This highlighted the dominant role of ATF4 in regulation of BiP, which is consistent with experimental evidence \cite{ma2003delineation}. Regulation of BiP was the critical regulator of spliced XBP1 (XBP1s), which in turn acts as a key marker of progression through different stages of UPR (supplementary materials Fig. S\ref{fg:Flux_Supp_Fig1}E). ATF4, cleaved ATF6 and XBP1s act as integrators of the signals coming from all the three branches of UPR and furthermore leads to regulation of BiP, thereby leading to a negative feedback or control of UPR signal. Another interesting note was the regulation of pro-apoptosis phenotype via regulation of Bcl2. PERK and ATF4 KO lead to delay in the onset of apoptosis (marked by slower and lower reduction of Bcl2 levels, supplementary materials Fig. S\ref{fg:Flux_Supp_Fig1}F). This effect could be attributed to the lack of CHOP mediated branch of Bcl2 regulation. On the other hand, IRE1 and CHOP KO leads to drastic reduction in apoptosis (marked by little or no change of Bcl2 levels, supplementary materials Fig. S\ref{fg:Flux_Supp_Fig1}F). CHOP KO, implicated the importance of CHOP in the down-regulation of Bcl2. IRE1 KO implicated the critical role of IRE1-TRAF2 mediated route of apoptosis. Overall signal flow analysis highlighted the extensive amount of crosstalk within the three branches of the UPR network. Taken together, the population of EUPR models recapitulated both short- and long timescale behavior following overload of unfolded proteins in a variety of cell types. It also captured the integration between multiple pathways and generated specific and testable hypothesis about the role of network components in signal propagation.

\subsection*{Sensitivity analysis stratified locally and globally important components of the UPR architecture.} 
First-order sensitivity coefficients were computed, time-averaged (over approximately 8 hrs of simulated time following ER stress) and rank-ordered for the 1090 models parameters and 636 species under normal and UPR induced conditions. Five parameter sets were selected from the parent ensemble based on rightful representation of the diversity for the sensitivity calculations (further details in the supplementary materials). We selected only a few parameter sets for the sensitivity analysis, because of the computational cost of computing all 1090 coefficients over several hours of simulated time. However, we did select diverse sets as shown by the scatter in the CV values (supplementary materials Fig. S\ref{fg:Para_S1}). Infrastructure parameters e.g. nuclear transport, RNA polymerase or ribosome binding were globally critical, independent of stress (black points, Fig. \ref{fg:fig5}). Additionally, apoptotic species and parameters were also important, both in the presence and absence of UPR (yellow points, Fig. \ref{fg:fig5}). Thus, as expected, components such as RNA polymerase or caspase activation were globally important irrespective of the folding state of the ER. More interesting however were coefficients that shifted above or below the 45$^{o}$-line in the presence of UPR. These points denote differentially important network components. While the majority of parameters and species became more important in the presence of stress, we found a band of parameters (Fig. \ref{fg:fig5} Inset) that were differentially important under stress. For example, the rank-ordering of the sensor and stress-transducer modules clearly increased in the presence of UPR. Approximately, 172 or 15\% of the parameters were significantly more important in the presence of UPR. These parameters were largely associated with adaptation and processing of unfolded or misfolded proteins, e.g., unfolded protein degradation, cleaved ATF6-induced gene expression, IRE1-TRAF2 mediated apoptosis regulation and RCAN1 regulation. Likewise, 75 or 12\% of the species were significantly more important in UPR compared with normal protein loads (data not shown).

Sensitivity analysis conducted over discrete two hour time windows revealed the time evolution of the importance of UPR network modules (Fig. \ref{fg:fig5}). Comparison of the 0 - 2 hrs time window with itself (top panel, first column of Fig. \ref{fg:fig5}), supported the earlier results that infrastructure components were globally critical followed by ERAD species. These species remained important in all time windows. On the other hand, during the initial 0 - 2 hrs window, ER stress transduction pathway components were robust. Comparison of the 0 - 2 hrs time window with later time points (working down the first column, Fig. \ref{fg:fig5}), showed the increasing importance of different modules as a function of time. For example, components of the PERK and IRE1 modules were more important in the 2 - 4 hrs window compared to the earlier time points, while alarm and apoptotic phase species were more important in the 6 - 8 hrs window compared to the earlier time points. Specifically, signal integration via the transcriptional activity of ATF6, ATF4 and XBP1s along with RCAN1 and cIAP role in apoptosis were significantly more important at 6-8 hrs as compared to 0-2 hrs time window. This is consistent with the dominant role of the negative feedback via the transcriptional regulation of BiP in UPR. Interestingly, the majority of species rankings were similar after 6 hrs (bottom row,  Fig. \ref{fg:fig5} and supplementary materials Fig. S\ref{fg:Sens_S1}). This analysis supports the structural analysis results, wherein the signal integrators like XBP1s and transcriptional regulation of BiP via ATF6 and ATF4 are the key regulators of the phased response of UPR.

To further investigate the role of positive BiP regulation via ATF6, ATF4 and XBP1s, we conducted sensitivity analysis upon knocking out the feedback branches of BiP for the nominal parameter set (supplementary materials Fig. S\ref{fg:Sens_S3}). Interestingly, upon knockout of any individual feedback branch like that of ATF4, ATF6 and XBP1s, the system overall remains equally robust. However the sensitivity of the alternate feedback components increases. This was most evident upon ATF4 feedback KO. We distinctly saw increase in sensitivity of feedback components associated with XBP1s and ATF6 (supplementary materials Fig. S\ref{fg:Sens_S3}). Upon ATF6 and XBP1s feedback KO, there wasn't much change in terms of sensitivity of the system. This further attests the key regulatory effect of ATF4 in mediating the positive BiP feedback which is an essential component of the adaptation phase of UPR. Another interesting observation was that when we completely knockout all the feedback branches of BiP in the adaptation phase, the system overall becomes relatively more robust (supplementary materials Fig. S\ref{fg:Sens_S3}). We distinctly saw a major shift of sensitivity of BiP upon removal of positive feedback. Overall $\sim$ 54 \% of the parameters were differentially less sensitive upon removal of BiP feedback as compared to WT. This brings to light how the presence of BiP feedback makes the system more susceptible/sensitive to perturbations. This highlights the essential role and relevance of targeting the BiP feedback in manipulating UPR and  specific importance of ATF4 amongst the three feedback branches.

\subsection*{Robustness analysis predicted the phenotypic consequence of structural perturbations to the UPR network.} We calculated the direction of unfolded protein load induced concentration shifts for 636 markers following single parameter knockouts (Edge KO), single gene knockouts (GKO) and single gene overexpression (GOX) to the UPR network for the nominal parameter set (Fig. \ref{fg:robustness_figure}). Robustness coefficients were used to quantify the effect of structural perturbations on network markers. Coefficients with values $>$ 1 ($<$ 1) indicated a marker increased (decreased) compared to the basal state, while a value $\sim$ 1 indicated approximately no change following a perturbation. Phenotypic behavior of the UPR models were analyzed as a result of perturbation, using downstream markers like Caspase 3 (marker for cell death), Bcl2 (marker for cell-survival) and other key signal integrators of the signals from ER stress transducers e.g., XBP1s, CHOP and ATF4.
